---
title: 'Maps: Exercises'
output: html_document
---

## Exercises {#ex}

These exercises accompany the [Maps](http://rpubs.com/NateByers/maps) tutorial. 

Use the following code to download 2014 daily No~2~ data for the United States. 
This data will be used in the excercises.

```{r, eval=FALSE}
# create a temporary file
temp <- tempfile()

# download the .zip file to a temporary file--this will take several minutes
download.file('http://aqsdr1.epa.gov/aqsweb/aqstmp/airdata/daily_42602_2014.zip', temp)

# unzip temporary file to your working directory
unzip(temp)

# delete the temporary file
unlink(temp)

# read the data into R
no2 <- read.csv('daily_42602_2014.csv', stringsAsFactors = FALSE)
```

```{r, echo=FALSE}
no2 <- read.csv('daily_42602_2014.csv', stringsAsFactors = FALSE)
```

1. Use `dplyr` to summarize the `no2` data frame so that each monitor has
one record with a column for the NO~2~ mean for the entire year and a column
for the standard deviation. 

[Solution](#ex1)

2. Use the data frame created in the first exercise to create a 
map of the monitors (use `map()` and `points()`).

[Solution](#ex2)

3. Use the data frame created in the first exercise and subset
it down to monitors in Ohio. Plot those monitors on a map and
label them with the "City.Name" column.

[Solution](#ex3)

4. Use `googleVis` to make a Google map of the Ohio monitors.

[Solution](#ex4)

5. Use `leaflet` to make a Leaflet map of the Ohio monitors.

[Solution](#ex5)

## Solutions

### Solution 1{#ex1}


```{r, warning=FALSE, message=FALSE}
library(dplyr)
as.tbl(no2)
# filter out days with less than 75% valid hours
no2 <- filter(no2, Observation.Percent >= 75)
no2_annual <- group_by(no2, State.Code, County.Code, Site.Num, 
                    Latitude, Longitude, Datum, State.Name, 
                    County.Name, City.Name, CBSA.Name)
no2_annual <- summarize(no2_annual, no2_mean = mean(Arithmetic.Mean, na.rm = TRUE),
                        no2_sd = sd(Arithmetic.Mean, na.rm = TRUE))
as.tbl(no2_annual)
```

[Back to exercises](#ex)

### Solution 2 {#ex2}

```{r, message=FALSE, warning=FALSE}
library(maps)
map(database = 'usa')
points(x = no2_annual$Longitude, y = no2_annual$Latitude,pch = 19)
```

[Back to exercises](#ex)

### Solution 3 {#ex3}

```{r}
ohio <- filter(no2_annual, State.Name == "Ohio")
map(database = 'state', regions = "ohio")
points(x = ohio$Longitude, y = ohio$Latitude,pch = 19)
title(main = "Ohio NO2 Monitor Locations")
text(x = ohio$Longitude, y = ohio$Latitude, labels = ohio$City.Name, pos = 4)
```


[Back to exercises](#ex)

### Solution 4 {#ex4}

```{r, warning=FALSE, message=FALSE}
library(googleVis)
```

```{r, echo=FALSE}
op <- options(gvis.plot.tag='chart')
```

```{r, results='asis', tidy=FALSE}
ohio$google.location <- paste(ohio$Latitude, ohio$Longitude, sep = ":")
# here I create a more informative label
ohio$label <- paste("NO2 Annual Mean:", round(ohio$no2_mean, 3))
g.inter.map <- gvisMap(data = ohio, locationvar = "google.location",
                       tipvar = "label")
plot(g.inter.map)
```

```{r, echo=FALSE}
options(op)
```


[Back to exercises](#ex)

### Solution 5 {#ex5}

```{r, warning=FALSE}
library(leaflet)
m <- leaflet()
m <- addTiles(m)
m <- addMarkers(m, lng=ohio$Longitude, lat=ohio$Latitude, 
                popup=ohio$label)
m
```


[Back to exercises](#ex)