---
title: "Maps"
output: html_document
---

In this tutorial we'll cover the following topics:

* [Basic Maps](#basic)
* [Google Maps](#google)
* [Leaflet Maps](#leaflet)


# Basic Maps {#basic}

There is no function for displaying maps in the base R functionality. To display
a simple map, we use the `maps` package. First, let's create a simple Indiana map.

```{r, warning=FALSE, message=FALSE}
library(maps)
map(database = 'state', regions = 'indiana')
```

Now we can add a few monitors as points. Use the longitude column
as the x-axis values and longitude for the y-axis values.

```{r}
monitors <- read.table(header=T, text='
  monitorid        lat        long   datum                    name            
          1   41.60668  -87.304729   WGS84              Gary-IITRI
          2  39.811097	-86.114469   WGS84  Indpls-Washington-Park
          3  39.749019	-86.186314   WGS84      Indpls-Harding-St.
          4  38.013248	-87.577856   WGS84  Evansville-Buena-Vista
          5  39.159383	-86.504762   WGS84             Bloomington
          6  39.997484  -86.395172   WGS84              Whitestown
                       ')
```

```{r, eval=FALSE}
points(x = monitors$long, y = monitors$lat)
```

```{r, echo=FALSE}
map(database = 'state', regions = 'indiana')
points(x = monitors$long, y = monitors$lat)
```

We can jazz it up a bit by making it a county map and changing the symbol type and
color.

```{r}
map(database = 'county', regions = 'indiana')
points(x = monitors$long, y = monitors$lat, pch = 19, col = 'red')
```

We can also select specific counties to look at, and give the map a title.

```{r}
map(database = 'county', regions = c('indiana,marion', 'indiana,boone'))
points(x = monitors$long, y = monitors$lat, pch = 19, col = 'red')
title(main = "Air Monitor Locations")
```

If we want the counties to be automatically labeled, we use the `map.text` 
function.

```{r}
map.text(database = 'county', regions = c('indiana,marion', 'indiana,boone'))
points(x = monitors$long, y = monitors$lat, pch = 19, col = 'red')
title(main = "Air Monitor Locations")
```


And if we want to label the monitors, we use the `text()` function.

```{r}
map(database = 'county', regions = c('indiana,marion', 'indiana,boone'))
points(x = monitors$long, y = monitors$lat, pch = 19, col = 'red')
title(main = "Air Monitor Locations")
text(x = monitors$long, y = monitors$lat, labels = monitors$name, pos = 2)
```



# Google Maps {#google}


To make interactive Google maps, we can use the `googleVis` package.

```{r, warning=FALSE, message=FALSE}
library(googleVis)
```

```{r, echo=FALSE}
op <- options(gvis.plot.tag='chart')
```

We have to do some formatting first. The coordinates must be in the form `lat:long`
so we need to create a data frame with a variable in that format.

```{r, results='asis', tidy=FALSE}
google.location <- paste(monitors$lat, monitors$long, sep = ":")
monitors.google <- data.frame(monitors, google.location)
g.inter.map <- gvisMap(data = monitors.google, locationvar = "google.location",
                       tipvar = "name")
plot(g.inter.map)
```

```{r, echo=FALSE}
options(op)
```


# Leaflet Maps {#leaflet}


Leaflet is another interactive map that can be created in R using the `leaflet`
package.

```{r, warning=FALSE}
library(leaflet)
m <- leaflet()
m <- addTiles(m)
m <- addMarkers(m, lng=monitors$long, lat=monitors$lat, 
                popup=monitors$name)
m
```



