---
title: "Maps and Spatial Data Analysis"
output: html_document
---

Many R packages have been developed to visualize and analyze spatial data. The best 
comprehensive resource that I've found on this topic is the book
[Applied Spatial Data Analysis with R](http://www.amazon.com/Applied-Spatial-Data-Analysis-Use/dp/1461476178). This tutorial relies heavily on that book.

- [The sp package and Spatial* objects](#sp)
- [Plotting spatial data](#plot)
- [Importing GIS data](#GIS)
- [Interpolation](#interpolation)

# The `sp` package and `Spatial*` objects {#sp}

The `sp` package is a foundational package for dealing with spatial data. It
provides classes and methods for making S4 objects in R. (For an explanation of 
S4 objects, see Hadley Wickham's page on object oriented programming in 
[Advanced R](http://adv-r.had.co.nz/OO-essentials.html)). The most basic is a `Spatial`
object. This object just contains information for a boundary box and the coordinate system.

## `Spatial`

Let's create a boundary box in the Chicago area. We'll use the constructor function
that has the same name as the class, `Spatial()`. The two arguments are `bbox` and
`proj4string`. The `bbox` argument is for the bounding box. It takes a matrix
with at least two rows, and the column names must be `min` and `max`. The first 
row contains the eastings (x-axis, or the longitude) and the second row contains
the northings (y-axis, or latitude). 


```{r}
# create a boundary box
longitudes <- c(-88.156775, -87.589771)
latitudes <- c(41.652208, 42.154143)
bounding_box <- matrix(c(longitudes, latitudes), nrow = 2, byrow = TRUE,
                       dimnames = list(NULL, c("min", "max")))
```

The second argument, `proj4string`, takes a string that specifies which projection
the coordinates are in. See the [`proj4` package](https://cran.r-project.org/web/packages/proj4/) 
for details on which strings to use. The simplest string is "+proj=longlat".

```{r}
projection <- "+proj=longlat" 
```

Now we can create the `Spatial` object. (The projection string must be wrapped in the 
`CRS()` function).

```{r}
library(sp)
chicago_sp <- Spatial(bbox = bounding_box, proj4string = CRS(projection))
chicago_sp
```

## `SpatialPoints`

A `Spatial` object isn't much good by itself, but it's the base object that other 
sub-classes are built on. To see the list of sub-classes, use the `getClass()` function.

```{r}
getClass("Spatial")
```

A sub-class is a class whose objects contain all of the information in the parent class
plus additional information. The most useful direct sub-class of `Spatial`, for our purposes,
is `SpatialPoints`. Typically we will want to look at monitor locations, so we need to 
represent those as points in a `SpatialPoints` object. We'll use the `airdata` dataset from
the `region5air` package.

```{r, message=FALSE}
library(region5air)
library(dplyr)
data(airdata)
as.tbl(airdata)
```