---
title: "Maps in R: Exercises"
output: 
  html_document: 
    theme: readable
---

These exercises accompany the Maps in R tutorial: 
<a href="http://rpubs.com/NateByers/Maps" target="_blank">
http://rpubs.com/NateByers/Maps</a>. These exercises use daily No2 data for the
United States in 2014. Run the following code to clean out your global
environment and load the data you need:

```{r, message=FALSE}
rm(list = ls())
library(dplyr)
library(region5air)
data(no2_2014)
```

To see where this data came from, see the [Download](#download) section below.

## Exercises {#ex}



[Solution 1](#ex1)



[Solution 2](#ex2)



[Solution 3](#ex3)

----------------------------------

#### Advanced Exercise



[Solution 4](#ex4)




--------------------------------------

## Solutions

#### Solution 1{#ex1}


[Back to exercises](#ex)

#### Solution 2{#ex2}



[Back to exercises](#ex)

#### Solution 3{#ex3}



[Back to exercises](#ex)

#### Solution 4{#ex4}



[Back to exercises](#ex)

## Download {#download}

Below is the code for retrieving the `no2_2014` data frame from the 
<a href="http://aqsdr1.epa.gov/aqsweb/aqstmp/airdata/download_files.html" target="_blank">
EPA website</a>.

```{r, eval=FALSE}
# create a temporary file
temp <- tempfile()

# download the .zip file to a temporary file--this might take a minute or two
download.file('http://aqsdr1.epa.gov/aqsweb/aqstmp/airdata/daily_42602_2014.zip', temp)

# unzip temporary file to your working directory
unzip(temp)

# delete the temporary file
unlink(temp)

# read the data into R
no2 <- read.csv('daily_42602_2014.csv', stringsAsFactors = FALSE)

# delete the .csv file from your working directory
unlink('daily_42602_2014.csv')
```

