---
title: 'Maps in R: Exercises'
output:
  html_document:
    theme: readable
  word_document: default
---

These exercises accompany the Maps in R tutorial: 
<a href="http://rpubs.com/NateByers/Maps" target="_blank">
http://rpubs.com/NateByers/Maps</a>. These exercises use daily No~2~ data for the
United States in 2014. Run the following code to clean out your global
environment and load the data you need:

```{r, message=FALSE}
rm(list = ls())
library(dplyr)
library(region5air)
data(no2_2014)
```

If you are curious about  where this data came from, see the [Download](#download)
section below.

## Exercises {#ex}

1. Use `dplyr` to create a data frame called `no2_max`. It should be a filtered
subset of the `no2_2014` data frame so that each monitor has one record with 
the highest 1-hour NO~2~ maximum for the entire year. __Hint:__ Take a look at 
the column names, `names(no2_2014)`. Use `group_by()` to group by state, county,
and site. Use `filter()` to find which 1st-max-value is equal to the maximum of
the 1st-max-value column (use `==`).

[Solution 1](#ex1)

2. Make a map of NO~2~ monitors. Use `map(database = 'usa')` to make a map of the
US first. Then use `points()` to add the monitors from the `no2_max` data frame.


[Solution 2](#ex2)

3. Use `filter()` to make a new data frame of NO~2~ motors that had a 1 hour max
value above 100 ppb. Call the new data frame `no2_violations` and map those monitors.

[Solution 3](#ex3)

4. Use the `leaflet` package to make an interactive map of the monitors in the
`no2_violations` data frame.



[Solution 4](#ex4)

----------------------------------

#### Advanced Exercise

5. Use the `choroplethr` package to make a heat map of the number of NO~2~ 
monitors in each state. __Hint:__ You will need to make a new data frame called
`no2_monitors`. Start with the `no2_max` data frame and group by state. Use the
`summarize()` function to get the total number of monitors in each group. The `n()`
function will count the total number of rows in a group.

[Solution 5](#ex5)




--------------------------------------

## Solutions

#### Solution 1{#ex1}

Here are the columns.

```{r}
names(no2_2014)
```

Now we know the column names for state, county, and site. Let's group by those
columns.

```{r}
no2_max <- group_by(no2_2014, State.Name, County.Name, Site.Num)
```

Now we'll filter the data frame so that the "X1st.Max.Value" is equivalent to
the maximum for each group.

```{r}
no2_max <- filter(no2_max, X1st.Max.Value == max(X1st.Max.Value, na.rm = TRUE))
```


[Back to exercises](#ex)

#### Solution 2{#ex2}

```{r, message=FALSE, warning=FALSE}
library(maps)
map(database = 'usa')
points(x = no2_max$Longitude, y = no2_max$Latitude, pch = 19)
```

[Back to exercises](#ex)

#### Solution 3{#ex3}

```{r}
no2_violations <- filter(no2_max, X1st.Max.Value > 100)
map(database = 'usa')
points(x = no2_violations$Longitude, y = no2_violations$Latitude, pch = 19,
       col = "red")
```


[Back to exercises](#ex)

#### Solution 4{#ex4}


```{r}
library(leaflet)
m <- leaflet()
m <- addTiles(m)
m <- addMarkers(m, lng=no2_violations$Longitude, lat=no2_violations$Latitude, 
                popup=no2_violations$CBSA.Name)
m
```


[Back to exercises](#ex)

#### Solution 5{#ex5}

```{r}
no2_monitors <- group_by(no2_max, State.Name)
no2_monitors <- summarize(no2_monitors, value = n())
```

The `choroplethr` package wants full state names with lower case. So we need to
remove Puerto Rico, turn the state values to all lower case, and rename the state
column to "region"

```{r}
no2_monitors <- filter(no2_monitors, State.Name != "Puerto Rico")
no2_monitors <- mutate(no2_monitors, State.Name = tolower(State.Name))
no2_monitors <- rename(no2_monitors, region = State.Name)
```

Now we can make the map.

```{r, message=FALSE, warning=FALSE}
library(choroplethr)
no2_map <- state_choropleth(no2_monitors, legend = "NO2 Monitors")
no2_map
```

Just for fun, let's see that with pipes.

```{r, warning=FALSE, message=FALSE}
no2_map_piped <- no2_max %>%
  group_by(State.Name) %>%
  summarize(value = n()) %>%
  filter(State.Name != "Puerto Rico") %>%
  mutate(State.Name = tolower(State.Name)) %>%
  rename(region = State.Name) %>%
  state_choropleth(legend = "NO2 Monitors")

no2_map_piped
```

[Back to exercises](#ex)

## Download {#download}

Below is the code for retrieving the `no2_2014` data frame from the 
<a href="http://aqsdr1.epa.gov/aqsweb/aqstmp/airdata/download_files.html" target="_blank">
EPA website</a>.

```{r, eval=FALSE}
# create a temporary file
temp <- tempfile()

# download the .zip file to a temporary file--this might take a minute or two
download.file('http://aqsdr1.epa.gov/aqsweb/aqstmp/airdata/daily_42602_2014.zip', temp)

# read the .csv file into R
no2 <- read.csv(unz(temp, 'daily_42602_2014.csv'), stringsAsFactors = FALSE)

# delete the temporary file
unlink(temp)
```

