---
title: "Maps"
output: html_document
---

In this tutorial we'll cover the following topics:

* [Basic Maps](#basic)
* [Heat Maps](#heat)
* [Google Maps](#google)
* [Leaflet Maps](#leaflet)


# Basic Maps {#basic}

There is no function for displaying maps in the base R functionality. To display
a simple map, we use the `maps` package. First, let's create a simple Indiana map.

```{r, warning=FALSE, message=FALSE}
library(maps)
map(database = 'state', regions = 'indiana')
```

Now we can add a few monitors as points. Use the longitude column
as the x-axis values and longitude for the y-axis values.

```{r}
monitors <- read.table(header=T, text='
  monitorid        lat        long   datum                    name            
          1   41.60668  -87.304729   WGS84              Gary-IITRI
          2  39.811097	-86.114469   WGS84  Indpls-Washington-Park
          3  39.749019	-86.186314   WGS84      Indpls-Harding-St.
          4  38.013248	-87.577856   WGS84  Evansville-Buena-Vista
          5  39.159383	-86.504762   WGS84             Bloomington
          6  39.997484  -86.395172   WGS84              Whitestown
                       ')
```

```{r, eval=FALSE}
points(x = monitors$long, y = monitors$lat)
```

```{r, echo=FALSE}
map(database = 'state', regions = 'indiana')
points(x = monitors$long, y = monitors$lat)
```

We can jazz it up a bit by making it a county map and changing the symbol type and
color.

```{r}
map(database = 'county', regions = 'indiana')
points(x = monitors$long, y = monitors$lat, pch = 19, col = 'red')
```

We can also select specific counties to look at, and give the map a title.

```{r}
map(database = 'county', regions = c('indiana,marion', 'indiana,boone'))
points(x = monitors$long, y = monitors$lat, pch = 19, col = 'red')
title(main = "Air Monitor Locations")
```

If we want the counties to be automatically labeled, we use the `map.text` 
function.

```{r}
map.text(database = 'county', regions = c('indiana,marion', 'indiana,boone'))
points(x = monitors$long, y = monitors$lat, pch = 19, col = 'red')
title(main = "Air Monitor Locations")
```


And if we want to label the monitors, we use the `text()` function.

```{r}
map(database = 'county', regions = c('indiana,marion', 'indiana,boone'))
points(x = monitors$long, y = monitors$lat, pch = 19, col = 'red')
title(main = "Air Monitor Locations")
text(x = monitors$long, y = monitors$lat, labels = monitors$name, pos = 2)
```

# Heat Maps {#heat}

The `choroplethr` packages makes it fairly easy to create heat maps in R. 
For illustration, we'll grab some data on non-attainment areas from 
<a href="http://www3.epa.gov/airquality/greenbook/hntc.html" target="_blank">this
EPA site</a>. 

```{r, eval=FALSE}
library(XML)
library(dplyr)

# scrape the table
table <- readHTMLTable("http://www3.epa.gov/airquality/greenbook/hntc.html")[[1]]

# get the state column as a vector
states <- table[, dim(table)[2]]

# change to character
states <- as.character(states)

# split the hyphenated elements 
states <- lapply(states, strsplit, split = "-")

# make one vector
nonattain <- c()
for(i in states){
  x <- i[[1]]
  if(!is.na(x[1])){
    nonattain <- c(nonattain, x)
  }
}

# make table of total nonattainment areas per state
nonattain_df <- data.frame(state = nonattain)
nonattain_df <- group_by(nonattain_df, state)
nonattain_df <- summarize(nonattain_df, value = n())

nonattain_df
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(XML)
library(dplyr)
table <- read.csv("nonattainment.csv", stringsAsFactors = FALSE)
# get the state column as a vector
states <- table[, dim(table)[2]]

# change to character
states <- as.character(states)

# split the hyphenated elements 
states <- lapply(states, strsplit, split = "-")

# make one vector
nonattain <- c()
for(i in states){
  x <- i[[1]]
  if(!is.na(x[1])){
    nonattain <- c(nonattain, x)
  }
}

# make table of total nonattainment areas per state
nonattain_df <- data.frame(state = nonattain)
nonattain_df <- group_by(nonattain_df, state)
nonattain_df <- summarize(nonattain_df, value = n())

nonattain_df
```

Now we'll use the `choroplethr` package to make a heat map
of the number of nonattainment areas for 8-hr ozone (2008 standard) by state.

```{r, eval=FALSE}
library(choroplethr)

# get state AQS table to get state names
state_codes <- read.csv("https://aqs.epa.gov/aqsweb/codes/data/StateCountyCodes.csv",
                        skip = 1, stringsAsFactors = FALSE)

state_codes <- unique(select(state_codes, State.Name, State.Abbr))

# full join with nonattain_df and convert NAs to zero
nonattain_df <- full_join(nonattain_df, state_codes, 
                          by = c("state" = "State.Abbr"))
nonattain_df[is.na(nonattain_df)] <- 0

# filter out regions that are not states and make lowercase state column
nonattain_df <- filter(nonattain_df, 
                       !state %in% c("CC", "MX", "GU", "PR", "VI"))
nonattain_df$region <- tolower(nonattain_df$State.Name)

nonattain_map <- state_choropleth(nonattain_df[, c("region", "value")], 
                                  title = "2008 8-hr Ozone Standard",
                                  legend = "Nonattain Areas")
nonattain_map
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=8}
library(choroplethr)

# get state AQS table to get state names
state_codes <- read.csv("https://aqs.epa.gov/aqsweb/codes/data/StateCountyCodes.csv",
                        skip = 1, stringsAsFactors = FALSE)

state_codes <- unique(select(state_codes, State.Name, State.Abbr))

# full join with nonattain_df and convert NAs to zero
nonattain_df <- full_join(nonattain_df, state_codes, 
                          by = c("state" = "State.Abbr"))
nonattain_df[is.na(nonattain_df)] <- 0

# filter out regions that are not states and make lowercase state column
nonattain_df <- filter(nonattain_df, 
                       !state %in% c("CC", "MX", "GU", "PR", "VI"))
nonattain_df$region <- tolower(nonattain_df$State.Name)

nonattain_map <- state_choropleth(nonattain_df[, c("region", "value")], 
                                  title = "2008 8-hr Ozone Standard",
                                  legend = "Nonattain Areas")
nonattain_map
```

# Google Maps {#google}


To make interactive Google maps, we can use the `googleVis` package.

```{r, warning=FALSE, message=FALSE}
library(googleVis)
```

```{r, echo=FALSE}
op <- options(gvis.plot.tag='chart')
```

We have to do some formatting first. The coordinates must be in the form `lat:long`
so we need to create a data frame with a variable in that format.

```{r, results='asis', tidy=FALSE}
google.location <- paste(monitors$lat, monitors$long, sep = ":")
monitors.google <- data.frame(monitors, google.location)
g.inter.map <- gvisMap(data = monitors.google, locationvar = "google.location",
                       tipvar = "name")
plot(g.inter.map)
```

```{r, echo=FALSE}
options(op)
```


# Leaflet Maps {#leaflet}


Leaflet is another interactive map that can be created in R using the `leaflet`
package.

```{r, warning=FALSE}
library(leaflet)
m <- leaflet()
m <- addTiles(m)
m <- addMarkers(m, lng=monitors$long, lat=monitors$lat, 
                popup=monitors$name)
m
```

# Exercises

Go to the <a href="http://rpubs.com/NateByers/mapexercises" target="_blank">exercises page</a> for this tutorial and test your comprehension of this material.

