---
title: 'Data Manipulation: Part 1'
output: 
  html_document: 
    theme: readable
---

This tutorial introduces the `dplyr` package for general purpose data manipulation.
`dplyr` is written by <a href="http://had.co.nz/" target="_blank">Hadley Wickham</a> and can be
installed from CRAN using the `install.packages()` function. The GitHub page for
the project is found here: 
<a href="https://github.com/hadley/dplyr" target="_blank">
https://github.com/hadley/dplyr</a>. It makes processing data frames easier and 
quicker.

This tutorial will cover the following functions in the `dplyr` package:

- [select()](#select), selecting columns

- [arrange()](#arrange), reordering rows

- [filter()](#filter), filtering rows

It also covers the following advanced topics:

- [What does `%in%` do?](#in)

- [dplyr and databases](#databases)


In the folder we provided you will also find a printed version of the RStudio
cheatsheet for data wrangling found here: 
<a href="https://www.rstudio.com/resources/cheatsheets/" target="_blank">
https://www.rstudio.com/resources/cheatsheets/</a>. All of the functions from 
`dplyr` that we cover in this course can be found on that cheatsheet.

The data we will be using for demonstrations in this tutorial can be found in the
`region5air` package. That package is not on CRAN, but can be found on GitHub:
<a href="https://github.com/NateByers/region5air" target="_blank">https://github.com/NateByers/region5air</a>.
If you have not installed this package, below is code that will install it using
the `install_github()` function from the `devtools` package.

```{r, eval=FALSE}
library(devtools)
install.github("NateByers/region5air")
```

Once the package has been installed, we need to library it and load the `airdata`
data frame into our global environment.

```{r, eval=FALSE}
library(region5air)
data(airdata)
head(airdata)
```

```{r, echo=FALSE}
library(region5air)
data(airdata)
head(airdata[, 1:5])
```

For information about this data set, type `?airdata`.

## Select columns with select() {#select}

Using base R functions, if we wanted to select a few columns from `airdata`, we 
would need to either use numbers or names inside of the `[ , ]` function. In
`dplyr` we simply use the `select()` function. 

As with all `dplyr` functions, the first parameter is the data frame that you want
to perform the operation on. In our case we want to select a few columns from the
`airdata` data frame, so it will look something like this: `select(airdata, ...)`.
But what are the column names? We use the `names` function to find out.

```{r}
names(airdata)
```

Suppose we wanted to create a data frame with just columns "site", "datetime",
"parameter", and "value". This is how we would use the `select()` function to do
that.

```{r, message=FALSE, warning=FALSE}
library(dplyr)
d <- select(airdata, site, datetime, parameter, value)
head(d)
```

As with all `dplyr` functions, the output is always a data frame (or data-frame-
like) object.

You can also select consective columns by separating the first column and the last
column with a colon. We've seen this before. We can use the colon to create a
consecute series of integers.

```{r}
1:3
```

We can use that short hand to select the first three columns of a data frame.

```{r}
b <- airdata[, 1:3]
head(b)
```

In `select()` we can pick a first column by name and a last column by name and 
separate them with a colon. 

```{r}
d <- select(airdata, site, datetime:value)
head(d)
```

## Arrange rows with arrange() {#arrange}

In `dplyr` ordering a data frame is very easy. We use the `arrange()` function.
As always, the data frame is the first parameter. Then we provide the columns by 
which to arrange the data frame. If we wanted to arrange`airdata` by first ordering
the rows by parameter (pollutant/weather variable) and then by site, we would do this.

```{r}
airdata <- arrange(airdata, parameter, site)
```

Again, the output of a `dplyr` function is always some modified version of the 
data frame that you supplied in the first parameter.

If we wanted to arrange the data frame by descending order, we simply wrap the 
column with the `desc()` function.

```{r}
airdata <- arrange(airdata, desc(parameter), datetime)
```

## Filter rows with filter() {#filter}

To filter a data frame we need to use a logical vector. For example, if we wanted 
to filter `airdata` by pollutant and POC, we would do something like this in base R.

```{r}
b <- airdata[airdata$parameter  == 44201 & airdata$poc == 1, ]

# show the dimensions of the data frame--
# the first number is the total number of rows, the second is the total number of columns
dim(b)
```

The result of `airdata$parameter == 44201 & airdata$poc == 1` is a logical vector
as long as the number of rows in `airdata`. You must put in those dollar signs 
and separate the two comparisons with `&`. With `dplyr` you can use the `filter()`
function without dollar signs and separate the logical comparisons with a comma,
which is equivalent to `&`. 

```{r}
d <- filter(airdata, parameter == 44201, poc == 1)
dim(d)
```

If you want to use two logical comparisons and combine them with an OR, you must
use the `|` symbol. Suppose we want to subset down to all records with parameter
6201 and with a poc of 1 _or_ 2. 

```{r}
d <- filter(airdata, parameter == 44201, poc == 1 | poc == 2)
dim(d)
```

For review, here is a table of logical operators in R.

Operator | Description
---------|------------
<	| less than
<=	| less than or equal to
>	| greater than
>=	| greater than or equal to
==	| exactly equal to
!=	| not equal to
x \| y |	x OR y
x & y	| x AND y
x %in% y | compare consecutive elements of x to all elements of y


## Additional comments on `dplyr` 

- Some functions don't appear to be that much easier to use than base R (like 
`select()` or `arrange()`). But `dplyr` provides a suite of functions 
_with the same syntax_ so that you can easily remember them. 

- `dplyr` is fast.

- You can use `dplyr` 
<a href="https://cran.r-project.org/web/packages/dplyr/vignettes/databases.html" target="_blank">
with databases</a>.

## Exercises

Exercises for this tutorial can be found here: 
<a href="http://rpubs.com/NateByers/DataManip1Exercises" target="_blank">
http://rpubs.com/NateByers/datamanip1exercises</a>.

-------------------

## Advanced Topics

### What does `%in%` do? {#in}

`%in%` is one of the more difficult logical operators to understand, but it's
extremely useful. Here's a demonstration:

```{r}
c("b", "c", "z") %in% c("a", "b", "c", "d", "e", "f")

```

The operator takes the first element of the vector on the left and checks to see
if the element is in the vector on the right. Since "b" is in the vector on the 
right, the first value of the returned vector is `TRUE`. "c" is also in the vector
on the right, so the second value in the returned vector is `TRUE`. But "z" is not
in the vector on the right. The third value is `FALSE`, and the length of the returned
vector equals the length of the vector on the left, with is 3.

Now let's switch the vectors around.

```{r}
c("a", "b", "c", "d", "e", "f") %in% c("b", "c", "z")
```

The length of the output vector is 6, equal to the length of the vector on the 
left. The output vector is `TRUE` for letters that are in the vector on the right 
and `FALSE` for the letters that are not in the vector on the right.


### `dplyr` and Databases {#databases}

You can use `dplyr` functions directly on databases. The functions will actually
translate what you are doing in R to an SQL statement and pass that statement to 
the database. This is convenient if you have a large database and small RAM. Instead
of pulling a large amount of data into RAM and then subsetting it, you can subset
in the database itself and only pull a smaller amount of data into R. 

`dplyr` has built-in support for SQLite, MySQL, MariaDB, and PostgreSQL. Let's build
another simple SQLite database for demonstration.

```{r, eval=FALSE}
# get the Cities table from the AQS codes site and replace . with _ in names
cities <- read.csv("https://aqs.epa.gov/aqsweb/codes/data/CityNamesByState.csv",
                   skip = 1, stringsAsFactors = FALSE)
names(cities) <- sub("\\.", "_", names(cities))
head(cities)
```

```{r, echo=FALSE}
load("data.rda")
head(cities)
```

```{r, message=FALSE}
# how big is this thing?
object.size(cities)

library(RSQLite)

# create an SQLite database and save connection in db_cities object
db_cities <- dbConnect(drv = SQLite(), dbname = "dplyrTest.sqlite")

# load the Cities table
dbWriteTable(db_cities, "Cities", cities)
```

Now we make a `dplyr` connection to the database and the Cities table.

```{r}
# dplyr database connection
dplyr_db <- src_sqlite("dplyrTest.sqlite")

# dplyr table connection
cities_tbl <- tbl(dplyr_db, "Cities")
```

Now we'll filter down to Michigan cities by using the `dplyr` function `filter()`.

```{r}
mich_cities <- filter(cities_tbl, State_Name == "Michigan")
head(mich_cities)
```

Let's see how much space this object reqires.

```{r}
object.size(mich_cities)
```

Now we'll remove the `dplyr` connections, disconnect from the database, and the
database from our working directory.

```{r}
rm(dplyr_db, cities_tbl)
dbDisconnect(db_cities)
unlink("dplyrTest.sqlite")
```