---
title: 'Data Manipulation 1: Exercises'
output: 
  html_document: 
    theme: readable
---

These exercises accompany the Data Manipulation Part 1 tutorial: 
<a href="http://rpubs.com/NateByers/DataManip1" target="_blank">
http://rpubs.com/NateByers/DataManip1</a>. These exercises use data frames
from the `region5air` library. Run the following code to clean out your global
environment and load the data you need:

```{r, message=FALSE}
rm(list = ls())
library(dplyr)
library(region5air)
data(airdata)
data(chicago_air)
```

## Exercises {#ex}

1. Use `select()` on the `airdata` data frame to create a `monitors` data 
frame with columns "site", "lat", "lon", and "GISDatum".

[Solution 1](#ex1)

2. Use `arrange()` on `airdata` to order it by site then by parameter then by 
datetime.

[Solution 2](#ex2)

3. Use `filter()` on `airdata` to create a `pm` data frame of PM~2.5~ measurements 
(AQS code 88101) from site 840180890022 with hourly values above 35 ug/m^3^. 
__Hint__: The "site" column is a character class and the "parameter"" and "value" 
columns have a numeric class. Use quotes around characters and unquoted numbers 
for numeric values.

[Solution 3](#ex3)

----------------------------------

#### Advanced Exercises

4. From `chicago_air`, create a data frame with readings between September 1 and 
September 30 where temperature values were at or above 90 degrees Fahrenheit.

[Solution 4](#ex4)

5. Use `filter()` and `%in%` to filter the `airdata` data frame down to just
ozone (44201) and PM~2.5~ (88101). Remember, the "parameter" column is a 
character class, so use quotes around the AQS parameter codes. 

[Solution 5](#ex5)

--------------------------------------

## Solutions

#### Solution 1{#ex1}

```{r}
monitors <- select(airdata, site, lat, lon, GISDatum)
```

This returns a very long data frame with many duplicate values. You can use the `distinct()` function
from `dplyr` to remove the duplicated rows.

```{r}
# look at the dimensions of the data frame
# the first number is the total number of rows, the second is the columns
dim(monitors)

# remove duplicates 
monitors <- distinct(monitors)

dim(monitors)
```

[Back to exercises](#ex)

#### Solution 2{#ex2}

```{r}
airdata <- arrange(airdata, site, parameter, datetime)
```


[Back to exercises](#ex)

#### Solution 3{#ex3}

```{r}
pm <- filter(airdata, parameter == 88101, site == "840181270024", value > 35)
```

[Back to exercises](#ex)

#### Solution 4{#ex4}

If we want to filter using date ranges, we need to make sure that date values are one of the
date classes. In the `chicago_air` the date column is the character class, not a date class.

```{r}
class(chicago_air$date)
```

We can covert it to the Date class.

```{r}
# no need to supply a format paramter--the date column is already in default format
chicago_air$date <- as.Date(chicago_air$date)
class(chicago_air$date)
```

Now we can filter using dates.

```{r}
filter(chicago_air, date >= as.Date("2013-09-01"), date <= as.Date("2013-09-30"),
       temp >= 90)
```

Or we can simply filter on the month of September,

```{r}
filter(chicago_air, month == 9, temp >= 90)
```


[Back to exercises](#ex)

#### Solution 5{#ex5}

```{r}
oz_pm <- filter(airdata, parameter %in% c("44201", "88101"))
```
[Back to exercises](#ex)