---
title: 'Data Manipulation Part 1: Exercises'
output: html_document
---

## Exercises {#ex}

These exercises accompany the [Data Manipulation 1](http://rpubs.com/NateByers/datamanip1) tutorial. 

1. From `airdata`, use `select()` to create a `monitors` data frame with columns "site", "lat", "lon", and "GISDatum".

[Solution](#ex1)

2. From `airdata`, use `arrange()` to order `airdata` by site then by parameter then by date/time.

[Solution](#ex2)

3. From `airdata`, use `filter()` to create a data frame of PM~2.5~ measurements (AQS code 88101) from site 840180890022 with hourly values above 35 ug/m^3^.

[Solution](#ex3)

4. From `chicago_air`, create a data frame with readings between July 1 and July 31 where temperature
values were at or above 90 degrees Fahrenheit.

[Solution](#ex4)

5. From `airdata`, find the daily maximum 1-hour value for all ozone monitors.

[Solution](#ex5)

## Solutions

### Solution 1{#ex1}
```{r, message=FALSE, warning=FALSE}
library(region5air)
library(dplyr)

data(airdata)
monitors <- select(airdata, site, lat, lon, GISDatum)
head(monitors)
```
Or equivalently,
```{r}
monitors <- select(airdata, site, lat:GISDatum)
```

This returns a very long data frame with many duplicate values. You can use the `distinct()` function
from `dplyr` to remove the duplicated rows.

```{r}
dim(monitors)
monitors <- distinct(monitors)
dim(monitors)
```

[Back to exercises](#ex)

### Solution 2 {#ex2}

```{r}
as.tbl(airdata)
airdata <- arrange(airdata, site, parameter, datetime)
as.tbl(airdata)
```

[Back to exercises](#ex)

### Solution 3 {#ex3}

If we look at the classes of the columns we can see that site is character but parameter and value are
numeric. 

```{r}
as.tbl(airdata)
```

So we filter site using a string (i.e. put quotes around the numerals) instead of a number, 
but for parameter and poc we use numeric values.

```{r}
my_pm <- filter(airdata, site == "840181270024", parameter == 88101, value > 35)
as.tbl(my_pm)
```

[Back to exercises](#ex)

### Solution 4 {#ex4}

If we want to filter using date ranges, we need to make sure that date values are one of the
date classes. In the `chicago_air` the date column is the character class, not a date class.

```{r}
data(chicago_air)
as.tbl(chicago_air)
```

We can covert it to the Date class.

```{r}
chicago_air$date <- as.Date(chicago_air$date)
```

Now we can filter using dates.

```{r}
filter(chicago_air, date >= as.Date("2013-07-01"), date <= as.Date("2013-07-30"),
       temp >= 90)
```

Or we can simply filter on the month of July,

```{r}
filter(chicago_air, month == 7, temp >= 90)
```

[Back to exercises](#ex)

### Solution 5 {#ex5}

First we filter down to ozone values and look at how many monitors there are and if there
is more than one poc value for any monitor.

```{r}
chicago_o3 <- filter(airdata, parameter == 44201)
table(select(chicago_o3, site, poc))
```

There are 23 sites and only one poc value per site, so we don't have to worry about more than one
value per site/hour. We can simply group by site and date.

```{r}
# first make date column
chicago_o3$date <- substr(chicago_o3$datetime, 1, 8)
chicago_o3 <- group_by(chicago_o3, site, date)
```

If we did have sites that had more than one poc per site/hour, we could filter on the poc by grouping
on the hour. We could do this for the whole `airdata` dataset this way:

```{r, eval=FALSE}
lowest_poc <- group_by(airdata, site, datetime, parameter)
lowest_poc <- filter(lowest_poc, poc == min(poc))
```

Now we can summarize the `chicago_o3` data frame by finding the maximum 1-hour value for
each day.

```{r}
chicago_o3 <- summarize(chicago_o3, o3_1hr_max = max(value, na.rm = TRUE))
as.tbl(chicago_o3)
```

[Back to exercises](#ex)