---
title: 'Intro To Shiny: Tutorial'
runtime: shiny
output: html_document
---

## Exercises {#ex}

These exercises accompany the [Intro to Shiny](http://rpubs.com/NateByers/IntroToShiny) tutorial. 

1. If you have not already, create a folder called `ChicagoApp` in your
working directory and place the `ui.R` and `server.R` scripts from the 
tutorial in that folder. Use `runApp()` to open the application.

[Solution](#ex1)

2. Add a table of the data to the bottom of the application. Use
`dataTableOutput()` in the `ui.R` script and `renderDataTable()`
in the `server.r` script.

[Solution](#ex2)

3. Add a Leaflet map to the application with a marker at the location 
of the monitor (latitude 42.14 and longitude -87.79923).

[Solution](#ex3)

4. 

[Solution](#ex4)

## Solutions

### Solution 1{#ex1}

```{r, eval=FALSE}
runApp("ChicagoApp")
```

```{r, echo=FALSE}
shinyAppDir(
  "MyApp",
  options=list(
    width="100%", height=550
  )
)
```

[Back to exercises](#ex)

### Solution 2 {#ex2}


```{r, eval=FALSE}
library(shiny)

# Define UI for application that plots time series
shinyUI(fluidPage(

  # Application title
  titlePanel("Chicago Air"),

  # Sidebar with a dropdown for selecting parameter
  sidebarLayout(
    sidebarPanel(
      selectInput("parameter",
                  "Select Parameter:",
                  "Ozone" = "ozone",
                  "Temperature" = "temp",
                  "Solar Radiation" = "solar")
    ),

    # Show a plot of the time series
    mainPanel(
      plotOutput("timePlot"),
      dataTableOutput("dataTable")
    )
  )
))
```

```{r, eval=FALSE}

library(shiny)
library(region5air)
data(chicago_air)
chicago_air$date <- as.Date(chicago_air$date)

  # Define server logic required to plot the time series
shinyServer(function(input, output) {
  
  output$timePlot <- renderPlot({
    ggplot(chicago_air, aes_string("date", input$parameter)) + geom_line() 
  })
  
  output$dataTable <- renderDataTable({
    chicago_air
  })
})
```

```{r, eval=FALSE}
runApp("ChicagoApp")
```

```{r, echo=FALSE}
shinyAppDir(
  "MyApp2",
  options=list(
    width="100%", height=1550
  )
)
```


[Back to exercises](#ex)

### Solution 3 {#ex3}

```{r, eval=FALSE}
library(shiny)

# Define UI for application that plots time series
shinyUI(fluidPage(

  # Application title
  titlePanel("Chicago Air"),

  # Sidebar with a dropdown for selecting parameter
  sidebarLayout(
    sidebarPanel(
      selectInput("parameter",
                  "Select Parameter:",
                  "Ozone" = "ozone",
                  "Temperature" = "temp",
                  "Solar Radiation" = "solar")
    ),

    # Show a plot of the time series
    mainPanel(
      plotOutput("timePlot"),
      dataTableOutput("dataTable")
    )
  )
))
```

```{r, eval=FALSE}

library(shiny)
library(region5air)
data(chicago_air)
chicago_air$date <- as.Date(chicago_air$date)

  # Define server logic required to plot the time series
shinyServer(function(input, output) {
  
  output$timePlot <- renderPlot({
    ggplot(chicago_air, aes_string("date", input$parameter)) + geom_line() 
  })
  
  output$dataTable <- renderDataTable({
    chicago_air
  })
})
```

```{r, eval=FALSE}
runApp("ChicagoApp")
```

```{r, echo=FALSE}
shinyAppDir(
  "MyApp3",
  options=list(
    width="100%", height=1550
  )
)
```


[Back to exercises](#ex)

### Solution 4 {#ex4}


[Back to exercises](#ex)