---
title: "Intro To Shiny: Exercises"
resource_files:
- MyApp/ui.R
- MyApp/server.R
- MyApp2/server.R
- MyApp2/ui.R
- MyApp3/server.R
- MyApp3/ui.R
- MyApp5/server.R
- MyApp5/ui.R
runtime: shiny
output: html_document
---



These exercises accompany the 
<a href="https://natebyers.shinyapps.io/IntroToShiny" target="_blank">Intro to 
Shiny</a> tutorial. 

## Exercises {#ex}

1. If you have not already, create a folder called `ChicagoApp` in your
working directory and place the `ui.R` and `server.R` scripts from the 
tutorial in that folder. Use `runApp()` to open the application.

[Solution](#ex1)

----------------------------------

### Advanced Exercises

2. Add a table of the data to the bottom of the application. Use `dataTableOutput()`
in the `ui.R` script and `renderDataTable()` in the `server.R` script. See 
<a href="http://shiny.rstudio.com/articles/datatables.html" target="_blank">this 
page</a> for explanation of what DataTables are, and see 
<a href="http://shiny.rstudio.com/gallery/basic-datatable.html" target="_blank">
this page</a> for a live example of how to use these functions in a Shiny app.


[Solution](#ex2)

3. Add a Leaflet map to the application with a marker at the location 
of the monitor (latitude 42.14 and longitude -87.79923). See
<a href="https://rstudio.github.io/leaflet/shiny.html" target="_blank">this page
</a> for a demonstration of how to put a Leaflet map in a Shiny app. Remember to
library the `leaflet` package in your `ui.R` and `server.R` scripts.

[Solution](#ex3)

4. Modify the `ggplot2` output with the `plotly` package. See the 
<a href="https://plot.ly/r/" target="_blank">project website</a> for more information
on `plotly`. See 
<a href="https://plot.ly/r/shiny-tutorial/" target="_blank">the documentation</a>
on how to use the `ggplotly()` function in a Shiny app. And change `geom_line()` 
to `geom_point()` for a cleaner image.

[Solution](#ex4)

## Solutions

### Solution 1{#ex1}

```{r, eval=FALSE}
runApp("ChicagoApp")
```

```{r, echo=FALSE}
shinyAppDir(
  "MyApp",
  options=list(
    width="100%", height=550
  )
)
```

[Back to exercises](#ex)

### Solution 2 {#ex2}

__ui.R__

```{r, eval=FALSE}
library(shiny)

# Define UI for application that plots time series
shinyUI(fluidPage(

  # Application title
  titlePanel("Chicago Air"),

  # Sidebar with a dropdown for selecting parameter
  sidebarLayout(
    sidebarPanel(
      selectInput("parameter",
                  "Select Parameter:",
                  c("Ozone" = "ozone",
                  "Temperature" = "temp",
                  "Solar Radiation" = "solar"))
    ),

    # Show a plot of the time series
    mainPanel(
      plotOutput("timePlot"),
      dataTableOutput("dataTable")
    )
  )
))
```

__server.R__

```{r, eval=FALSE}

library(shiny)
library(region5air)
data(chicago_air)
chicago_air$date <- as.Date(chicago_air$date)

  # Define server logic required to plot the time series
shinyServer(function(input, output) {
  
  output$timePlot <- renderPlot({
    ggplot(chicago_air, aes_string("date", input$parameter)) + geom_line() 
  })
  
  output$dataTable <- renderDataTable({
    chicago_air
  })
})
```

```{r, eval=FALSE}
runApp("ChicagoApp")
```

```{r, echo=FALSE}
shinyAppDir(
  "MyApp2",
  options=list(
    width="100%", height=1600
  )
)
```


[Back to exercises](#ex)

### Solution 3 {#ex3}

__ui.R__

```{r, eval=FALSE}
library(shiny)
library(leaflet)

# Define UI for application that plots time series
shinyUI(fluidPage(

  # Application title
  titlePanel("Chicago Air"),

  # Sidebar with a dropdown for selecting parameter
  sidebarLayout(
    sidebarPanel(
      leafletOutput("mymap"),
      selectInput("parameter",
                  "Select Parameter:",
                  c("Ozone" = "ozone",
                  "Temperature" = "temp",
                  "Solar Radiation" = "solar"))
    ),

    # Show a plot of the time series
    mainPanel(
      plotOutput("timePlot"),
      dataTableOutput("dataTable")
    )
  )
))
```

__server.R__

```{r, eval=FALSE}

library(shiny)
library(region5air)
library(leaflet)
data(chicago_air)
chicago_air$date <- as.Date(chicago_air$date)

  # Define server logic required to plot the time series
shinyServer(function(input, output) {
  
  output$timePlot <- renderPlot({
    ggplot(chicago_air, aes_string("date", input$parameter)) + geom_line() 
  })
  
  output$dataTable <- renderDataTable({
    chicago_air
  })
  
  output$mymap <- renderLeaflet({
    library(leaflet)
    m <- leaflet()
    m <- addTiles(m)
    m <- addMarkers(m, lng=-87.79923, lat=42.14)
    setView(m, lng=-87.79923, lat=42.14, zoom = 10)
  })
})
```

```{r, eval=FALSE}
runApp("ChicagoApp")
```

```{r, echo=FALSE}
shinyAppDir(
  "MyApp3",
  options=list(
    width="100%", height=1600
  )
)
```


[Back to exercises](#ex)

### Solution 4 {#ex4}

__ui.R__

```{r, eval=FALSE}
library(shiny)
library(leaflet)
library(plotly)

# Define UI for application that plots time series
shinyUI(fluidPage(

  # Application title
  titlePanel("Chicago Air"),

  # Sidebar with a dropdown for selecting parameter
  sidebarLayout(
    sidebarPanel(
      leafletOutput("mymap"),
      selectInput("parameter",
                  "Select Parameter:",
                  c("Ozone" = "ozone",
                  "Temperature" = "temp",
                  "Solar Radiation" = "solar"))
    ),

    # Show a plot of the time series
    mainPanel(
      plotlyOutput("timePlot"),
      dataTableOutput("dataTable")
    )
  )
))
```

__server.R__

```{r, eval=FALSE}
library(shiny)
library(region5air)
library(leaflet)
library(ggplot2)
library(plotly)

data(chicago_air)
chicago_air$date <- as.Date(chicago_air$date)

shinyServer(function(input, output) {
  
  output$timePlot <- renderPlotly({
    g <- ggplot(chicago_air, aes_string("date", input$parameter)) + geom_point()
    ggplotly(g)
  })
  
  output$dataTable <- renderDataTable({
    chicago_air
  })
  
  output$mymap <- renderLeaflet({
    library(leaflet)
    m <- leaflet()
    m <- addTiles(m)
    m <- addMarkers(m, lng=-87.79923, lat=42.14)
    setView(m, lng=-87.79923, lat=42.14, zoom = 10)
  })
  
})
```

```{r, eval=FALSE}
runApp("ChicagoApp")
```

```{r, echo=FALSE}
shinyAppDir(
  "MyApp5",
  options=list(
    width="100%", height=1600
  )
)
```


[Back to exercises](#ex)