---
title: 'Reshaping Data: Exercises'
output:
  html_document:
    theme: readable
  word_document: default
---

These exercises accompany the Reshaping Data tutorial: 
<a href="http://rpubs.com/NateByers/Reshaping" target="_blank">
http://rpubs.com/NateByers/Reshaping</a>. The exercises use data frames
from the `region5air` library. Run the following code to clean out your global
environment and load the data you need:

```{r, message=FALSE}
rm(list = ls())
library(tidyr)
library(dplyr)
library(region5air)
data(airdata)
data(chicago_air)
```

## Exercises {#ex}

1. The `chicago_air` data frame is in a wide format. Use `gather()` to make a long
data frame named `chicago_air_long`. 

[Solution 1](#ex1)

2. The `airdata` data frame is in a long format. Use the `filter()` function to
create a data frame called `site22`. Filter down to site "840180890022" and a
poc of 1 (remember to use `==`). Use the `select()` function to select only the
"datetime", "parameter", and "value" columns. Use `spread()` on `site22` to make
a wide data frame called `site22_wide` with separate columns for each parameter.
__Hint:__ you want to spread the "parameter" column, so identify that column as
the `key` in the `spread()` function. The "value" column should be identified as
the `value` in the function.

[Solution 2](#ex2)

3. Use the `filter()` function on `airdata` to create a data frame called `pm25`.
Filter down to parameter "88101". Use the select() function to select only the 
“datetime”, “site”, and “value” columns. Use `spread()` on `pm25` to make a wide 
data frame called `pm25_wide` with separate columns for each site. __Hint:__ you
want to spread the "site" column, so identify that column as the `key` in the
`spread()` function.

[Solution 3](#ex3)

----------------------------------

#### Advanced Exercises

4. Use `ggplot2` to plot the `chicago_air_long` data frame that was created in 
exercise 1. First make sure to convert the "date" column to a `Date` class using
`as.Date()`. Use `facet_grid()` in the plot to make separate facets for each
parameter, and be sure to set the scales to "free".

[Solution 4](#ex4)




--------------------------------------

## Solutions

#### Solution 1{#ex1}

```{r}
chicago_air_long <- gather(chicago_air, key = "parameter", value = "value", 
                           ozone:solar)
```

[Back to exercises](#ex)

#### Solution 2{#ex2}

```{r}
site22 <- filter(airdata, site == "840180890022", poc == 1)
site22 <- select(site22, datetime, parameter, value)
site22_wide <- spread(site22, key = "parameter", value = "value")
```

[Back to exercises](#ex)

#### Solution 3{#ex3}

```{r}
pm25 <- filter(airdata, parameter == "88101")
pm25 <- select(pm25, datetime, site, value)
pm25_wide <- spread(pm25, key = "site", value = "value")
```


[Back to exercises](#ex)

#### Solution 4{#ex4}

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
chicago_air_long$date <- as.Date(chicago_air_long$date)
ggplot(chicago_air_long, aes(date, value)) + 
  geom_point() + facet_grid(parameter ~ ., scales = "free")
```

[Back to exercises](#ex)

