---
title: 'Merging: Exercises'
output: html_document
---

## Exercises {#ex}

These exercises accompany the 
<a href="http://rpubs.com/NateByers/merging" target="_blank">Merging</a> tutorial. 

1. Use the `merge()` function to merge `airdata` with the AQS units table found at 
https://aqs.epa.gov/aqsweb/codes/data/Units.csv.

[Solution](#ex1)

2. Do the same merge described in Exercise 1 but use the `dplyr` function `inner_join()`.

[Solution](#ex2)

3. Subset the AQS State and County code table (https://aqs.epa.gov/aqsweb/codes/data/StateCountyCodes.csv) 
down to just the counties that have monitors in `airdata` using the
`semi_join()` function.

[Solution](#ex3)

4. Subset the AQS Criteria table (https://aqs.epa.gov/aqsweb/codes/data/Parameters-CRITERIA.csv)
down to parameters that are *not* in `airdata` using the `anti_join()` function.

[Solution](#ex4)

5. Do a full merge of the `pollutants` and `monitors` data frames from the 
<a href="http://rpubs.com/NateByers/merging" target="_blank">Merging</a> tutorial. 
Do a semi join with the table below, but only keep records where the Violation 
column is TRUE.

```{r}
criteria <- read.table(header=T, text='
  Parameter   Duration    Monitor   Violation              
      ozone         1h          1       FALSE
        so2         1h          1        TRUE
      ozone         8h          2        TRUE
        no2         1h          4       FALSE
                       ')
```

[Solution](#ex5)

## Solutions

### Solution 1{#ex1}

```{r, warning=FALSE, message=FALSE}
library(region5air)
library(dplyr)
data(airdata)
as.tbl(airdata)
```

```{r, eval=FALSE}
aqs_units <- read.csv("https://aqs.epa.gov/aqsweb/codes/data/Units.csv",
                      skip = 1, stringsAsFactors = FALSE)
head(aqs_units)
```

```{r, echo=FALSE}
aqs_units <- read.csv("units.csv", stringsAsFactors = FALSE)
head(aqs_units)
```

```{r}
air <- merge(airdata, aqs_units, by.x = "unit", by.y = "Unit")
as.tbl(air)
```

[Back to exercises](#ex)

### Solution 2 {#ex2}

```{r, message=FALSE, warning=FALSE}
air <- inner_join(airdata, aqs_units, by = c("unit" = "Unit"))
as.tbl(air)
```

[Back to exercises](#ex)

### Solution 3 {#ex3}

```{r, eval=FALSE}
aqs_counties <- read.csv("https://aqs.epa.gov/aqsweb/codes/data/StateCountyCodes.csv",
                       skip = 1, stringsAsFactors = FALSE)
as.tbl(aqs_counties)
```

```{r, echo=FALSE}
aqs_counties <- read.csv("states.csv", stringsAsFactors = FALSE)
as.tbl(aqs_counties)
```

```{r}
airdata$State.Code <- substr(airdata$site, 4, 5)
airdata$County.Code <- as.integer(substr(airdata$site, 6, 8))
my_counties <- semi_join(aqs_counties, airdata)
my_counties
```

[Back to exercises](#ex)

### Solution 4 {#ex4}

First we read in the AQS table and narrow it down to just a few unique columns

```{r, eval=FALSE}
aqs_criteria <- read.csv("https://aqs.epa.gov/aqsweb/codes/data/Parameters-CRITERIA.csv",
                       skip = 1, stringsAsFactors = FALSE)
as.tbl(aqs_criteria)
```

```{r, echo=FALSE}
aqs_criteria <- read.csv("criteria.csv", stringsAsFactors = FALSE)
as.tbl(aqs_criteria)
```

```{r}
aqs_criteria <- distinct(select(aqs_criteria, Parameter.Code, 
                                Parameter.Desc, Duration.Desc))
as.tbl(aqs_criteria)
```

Now we filter down to rows with parameters that are not in `airdata`.

```{r}
missing_criteria <- anti_join(aqs_criteria, airdata, 
                              by = c("Parameter.Code" = "parameter"))
missing_criteria
```

[Back to exercises](#ex)

### Solution 5 {#ex5}

```{r}
monitors <- read.table(header=T, text='
  monitorid        lat       long                 
          1  42.467573 -87.810047     
          2  42.049148 -88.273029
          3  39.110539 -90.324080
                       ')

pollutants <- read.table(header=T, text='
  pollutant   duration    monitorid                 
      ozone         1h            1
        so2         1h            1
      ozone         8h            2
        no2         1h            4
                       ')

mon_pol <- full_join(monitors, pollutants)

criteria <- filter(criteria, Violation == TRUE)

mon_pol <- semi_join(mon_pol, criteria, 
                     by = c("pollutant" = "Parameter",
                            "duration" = "Duration",
                            "monitorid" = "Monitor"))

mon_pol
```

[Back to exercises](#ex)