---
title: "Merging in R"
output: 
  html_document: 
    theme: readable
---

This tutorial will cover how to merge data frames using `dplyr`. 

- [Mutating joins](#mutate)

- [Filtering joins](#filter)

- [A note on keys](#keys)

Advanced topic:

- [Joining database tables with dplyr](#database)

## Mutating joins {#mutate}

`dplyr` contains joining functions that enable you to combine two data frames
into one. These functions mimic database joins. If you aren't familiar with database 
joins, <a href="http://www.sql-join.com/" target="_blank">this site</a> has a good
overview.

We'll use two simple data frames to demonstrate how joins work.

```{r}
monitors <- read.table(header=T, text='
  monitorid        lat       long                 
          1  42.467573  -87.810047     
          2  42.049148  -88.273029
          3  39.110539  -90.324080
                       ')

pollutants <- read.table(header=T, text='
  pollutant   duration    monitorid                 
      ozone         1h            1
        so2         1h            1
      ozone         8h            2
        no2         1h            4
                         ')
```

To merge two data frames, there must be one or more columns that are in both data
frames. If we look at the `monitors` data frame and the `pollutants` data frame, 
we can see that they both have a "monitorid" column. That's the column we would
merge on, and in this tutorial we will refer to it as the "key".  

Below is a list of merging functions in `dplyr` that combine two data frames into
one, also referred to as mutating joins.

#### `inner_join()` 

This function combines the columns from both data frames, but only keeps rows where 
the value in the key column matches in both data frames. Below we do an inner 
join on our two data frames.

```{r, message=FALSE, warning=FALSE}
library(dplyr)

inner_join(pollutants, monitors, by = "monitorid")
```

Note that there is only one record for monitor 1 in the `monitors` data frame,
but the information for that monitor is repeated in the output because there is
more than one monitor id with the value 1 in the `pollutants` data frame.

#### `left_join()`

If we wanted to keep all of the pollutant information, even if there wasn't a 
corresponding record in the `monitors` data frame, then we could use the 
`left_join()` function.

```{r}
left_join(pollutants, monitors, by = "monitorid")
```

#### `full_join()`

We can also keep all records from both data frames using the `full_join()` 
function.


```{r}
full_join(pollutants, monitors, by = "monitorid")
```


## Filtering joins with `dplyr`{#filter}

Sometimes we would like to subset, or filter, a data frame based on the contents of
another data frame. In this kind of task we aren't combining two data frames, so
it's not a mutating join. We are filtering, so these are called filtering joins.

#### `semi_join()`

Suppose we wanted to keep the `pollutants` data frame in the same
format, but only keep the records that also have a match in the `monitors` data
frame. We can use the `semi_join()` function to filter the `pollutants` data
frame.

```{r}
semi_join(pollutants, monitors, by = "monitorid")
```

#### `anti_join()`

We may also want to know what records are in a data frame that do _not_ match the
records in another data frame. Suppose we want to know which monitors from the 
`monitors` data frame that are not also in the `pollutants` data frame. We can use
the `anti_join()` function.

```{r}
anti_join(monitors, pollutants, by = "monitorid")
```

## A note on keys {#keys}

The `dplyr` merging functions will merge two data frames on multiple columns with
differing names. Below we create two data frames for illustration.

```{r}
sites <- read.table(header=T, text='
  state   site         lat       long                 
     IL      1   42.467573  -87.810047     
     IL      2   42.049148  -88.273029
     IN      1   41.68140   -87.49473
                       ')

parameters <- read.table(header=T, text='
  region   monitor   parameter   duration                     
      IL         1       ozone         1h
      IN         1         so2         1h            
      IL         2       ozone         8h            
      IL         2         no2         1h            
                         ')

```

These two data frames can only be merged if we join on two keys, the column with
state information and the column with site information. Those two columns have
different names in the two data frames. We can do a full join of these two data 
frames if we supply the names of the keys in the `by` parameter. 

```{r}
full_join(sites, parameters, by = c("state" = "region", "site" = "monitor"))
```

## Exercises

Go to the <a href="http://rpubs.com/NateByers/MergingExercises" target="_blank">
exercises page</a> for this tutorial and test your comprehension of the material.

--------------------------------------------------

## Advanced Topic

### Joining database tables with `dplyr` {#database}

In <a href="http://rpubs.com/NateByers/DataManip1" target="_blank">another 
tutorial</a> we learned how to access a database table and manipulate it using
`dplyr` functions. You can also join two database tables within the database 
itself using `dplyr` functions. The functions translate the operation you want to
perform into a SQL expression. Then the expression gets passed to the database and
the result is returned to your R environment in a data-frame-like object.
