---
title: "Merging"
output: html_document
---

This turial will cover how to merge data frames using base R and `dplyr`. 

- [Base R `merge()`](#base)
- [Merging with `dplyr`](#dplyr)
- [Filtering joins with `dplyr`](#filter)

We will be using AQS tables to merge with a summarized version of the `airdata` dataset
from the `region5air` package.

```{r, message=FALSE, warning=FALSE}
library(region5air)
library(dplyr)

data(airdata)
airdata$year <- substr(airdata$datetime, start = 1, stop = 4)
airdata$state <- substr(airdata$site, start = 4, stop = 5)
air <- group_by(airdata, site, state, year, parameter, unit)
air <- summarise(air, mean = mean(value, na.rm = TRUE))
air
```

```{r, eval=FALSE}
aqs_states <- read.csv("https://aqs.epa.gov/aqsweb/codes/data/StateCountyCodes.csv",
                       skip = 1, stringsAsFactors = FALSE)
head(aqs_states)
```

```{r, echo=FALSE}
aqs_states <- read.csv("states.csv", stringsAsFactors = FALSE)
head(aqs_states)
```

# Merging in base R {#base}

Merging two data frames can be accomplished in base R using the `merge()` function. 
To merge, there must be one or more columns that the two tables have in common. If we 
wanted to merge the `air` dataset and the `aqs_states` table, we can see that the common
columns are "state" and "State.Code" respectively.

The first two arguments of the `merge()` function are the two data frames to be merged. 
If there are column names that are the same in the two data frames, then the default 
behavior is to the merge on those column names, and no other parameter is required.

As an example, let's make a copy of the `aqs_states` table and change the `State.Code`
name to `state`.

```{r}
b_states <- aqs_states
names(b_states)[names(b_states) == "State.Code"] <- "state"
```

Now we can merge by using just the two data frames as arguments.

```{r}
b <- merge(air, b_states)
head(as.tbl(b))
```

If there are no common names, then we must use the `by.x` and `by.y` parameters. 

```{r}
b <- merge(air, aqs_states, by.x = "state", by.y = "State.Code")
head(as.tbl(b))
```

The default operation of `merge()` is an inner join. This means that, in column(s) that
are shared among the two data frames, only records that match will be kept. In the example
above, only states with the state codes in the `aqs_states` data frame that were also in
the "state" column of the `air` data frame were kept.

```{r}
unique(air$state)
unique(aqs_states$State.Code)
unique(b$state)
```

You can also specify a left join or a right join if you would like to keep all of the
records from one of the data frames. To more easily see how this works, we'll create two
small data frames.

```{r}
monitors <- read.table(header=T, text='
  monitorid        lat       long                 
          1  42.467573 -87.810047     
          2  42.049148 -88.273029
          3  39.110539 -90.324080
                       ')

pollutants <- read.table(header=T, text='
  pollutant   duration    monitorid                 
      ozone         1h            1
        so2         1h            1
      ozone         8h            2
        no2         1h            4
                       ')
```

An inner join would keep only the monitors that are in common, which are 1 and 2.

```{r}
merge(pollutants, monitors)
```

Note that there is only one record for monitor 1 in the `monitors` data frame,
but the information for that monitor is repeated in the output because there is more
than one monitor id with the value 1 in the `pollutants` data frame.

If we wanted to keep all of the pollutant information, even if there wasn't a 
corresponding record in the `monitors` data frame, then we could do a left join
by using the `all.x` parameter.

```{r}
merge(pollutants, monitors, all.x = TRUE)
```

We can also keep all records from both data frames, which is called a full join. We
use the `all` parameter.

```{r}
merge(pollutants, monitors, all = TRUE)
```

# Merging with `dplyr` {#dplyr}

The `dplyr` package has functions that can do all of the operations that `merge()` can.
Similar to `merge()`, the functions in `dplyr` have you specify which colums to merge on,
or it merges on any column names that are in common. However, there is one `by` parameter.

```{r}
b <- inner_join(air, aqs_states, by = c("state" = "State.Code"))
head(b)
```

- `inner_join` Merge records that have shared values in the columns being merged on.

```{r}
inner_join(pollutants, monitors)
```

- `left_join` Keep all records in the first data frame.

```{r}
left_join(pollutants, monitors)
```

- `full_join` Keep all records in both data frames.

```{r}
full_join(pollutants, monitors)
```

# Filtering joins with `dplyr`{#filter}

Sometimes we would like to subset, or filter, a data frame based on the contents of
another data frame. Suppose we wanted to keep the `pollutants` data frame in the same
format, but only keep the records that also have a match in the `monitors` data
frame. We can use the `semi_join()` function from the `dplyr` package.

```{r}
semi_join(pollutants, monitors)
```

We may also want to know what records are in a data frame that do _not_ match the records
in another data frame. Suppose we want to know which monitors from the `monitors` data
frame that are not also in the `pollutants` data frame. We can use the `anti_join` function.

```{r}
anti_join(monitors, pollutants)
```
