---
title: 'Getting Data: Exercises'
output: 
  html_document: 
    theme: readable
---

## Exercises {#ex}

These exercises accompany the Getting Data tutorial: 
<a href="http://rpubs.com/NateByers/gettingData" target="_blank">
http://rpubs.com/NateByers/gettingData</a>. 

1. To use the `raqdm` package you must have a user name and password from EPA.
Register for an account here: <a href="https://aqs.epa.gov/api" target="_blank">
https://aqs.epa.gov/api</a>. Assuming you had a user name and account, write
R code that would request ozone data from all Illinois monitors between April 1 and
September 30, 2015. For a list of parameter and state codes, go to the AQS codes
website: <a href="http://www.epa.gov/aqs/aqs-code-list" target="_blank">
http://www.epa.gov/aqs/aqs-code-list</a>.

[Solution 1](#ex1)

2. Write code that uses `raqdm` to send and receive a request for all criteria
pollutant data for Cook County, Illinois between June 1 and August 31, 2015.
__Hint__: You can use a `pc` parameter in `getAQDMdata` that will take a text value
indicating a group of paramters ("CRITERIA"). You can also find county codes in
the AQS codes website:
<a href="http://www.epa.gov/aqs/aqs-code-list" target="_blank">
http://www.epa.gov/aqs/aqs-code-list</a>.

[Solution 2](#ex2)

3. Use the `read.csv()` function to create a data frame of unit measurement
information from the AQS codes website: <a href="http://www.epa.gov/aqs/aqs-code-list" target="_blank">
http://www.epa.gov/aqs/aqs-code-list</a>.

[Solution 3](#ex3)

#### Advanced Exercise

4. Create an `SQLite` database. Read in the parameters and units data from the 
AQS codes website to load the database with two tables: Parameters and Units.
Query the database and pull out a single table with parameters and unit information
in it. __Hint__: In your SQL statement, you will need to use a JOIN statement 
joining on the "Standard Unit" column in the Paramters table and the "Unit" column 
in the Unit table.

[Solution 4](#ex4)

## Solutions

#### Solution 1{#ex1}

First, library the package and set your user name and password if you have not
done so.
```{r eval=FALSE, message=FALSE, warning=FALSE}
library(raqdm)
setAQDMuser(user = "me@email.com", pw = "secret", save = TRUE)
```

On the AQS codes website we can find the number for Illinois (17) and the number
for ozone (44201). So we send a request using the `getAQDMdata()` function, assigning
the `synchronous` parameter a value of `FALSE`.

```{r, eval=FALSE}
ozone_req <- getAQDMdata(state = "17", bdate = "20150401", edate = "20150930", 
                         param = "44201", synchronous = FALSE)
```

Then wait for the email notification. Once it arrives, run the `getAQDMrequest()`
function.

```{r, eval=FALSE}
ozone <- getAQDMrequest(ozone_req)
head(ozone)
```
```{r, echo=FALSE}
load("exercise_data.rda")
ozone
```

[Back to exercises](#ex)

#### Solution 2{#ex2}

On the AQS codes website we can find the number for Cook County (031). We send a 
request using the `getAQDMdata()` function, assigning the `synchronous` parameter 
a value of `FALSE` and assigning the `pc` parameter a value of `"CRITERIA"`.

```{r, eval=FALSE}
criteria_req <- getAQDMdata(state = "17", county = "031", bdate = "20150601", 
                            edate = "20150831", pc = "CRITERIA", 
                            synchronous = FALSE)
```

Then wait for the email notification. Once it arrives, run the `getAQDMrequest()`
function.

```{r, eval=FALSE}
criteria <- getAQDMrequest(criteria_req)
head(criteria)
```
```{r, echo=FALSE}
criteria
```

[Back to exercises](#ex)

#### Solution 3{#ex3}

On the AQS codes website we can find the url for units by clicking on "Units"
and then clicking on "Download Delimited Version of the Code Table". Copy and paste
the url into `read.csv()` and assign the `skip` parameter a value of `1`.


```{r, eval=FALSE}
aqs_units <- read.csv("https://aqs.epa.gov/aqsweb/codes/data/Units.csv",
                      skip = 1, stringsAsFactors = FALSE)
head(aqs_units)
```
```{r, echo=FALSE}
head(aqs_units)
```

[Back to exercises](#ex)

#### Solution 4{#ex4}

Read the Parameters and Units into R from the AQS code website, if you have not
done so already.


```{r, eval=FALSE}
aqs_params <- read.csv("https://aqs.epa.gov/aqsweb/codes/data/ParametersByDesc.csv",
                       stringsAsFactors = FALSE, skip = 1)
aqs_units <- read.csv("https://aqs.epa.gov/aqsweb/codes/data/Units.csv",
                      skip = 1, stringsAsFactors = FALSE)
```

You will need to replace the `.` between the words in the column names with `_`,
or you won't be able to refer to field names in your SQL statement.

```{r, eval=FALSE}
names(aqs_params) <- sub("\\.", "_", names(aqs_params))
names(aqs_units) <- sub("\\.", "_", names(aqs_units))
```

Create the database in your working directory and load the two tables.

```{r}
library(RSQLite)
db_par_unit <- dbConnect(SQLite(), "par_unit.sqlite")
dbWriteTable(db_par_unit, "Parameters", aqs_params)
dbWriteTable(db_par_unit, "Units", aqs_units)
dbListTables(db_par_unit)
```

Now we write a SQL statement that joins the "Standard_Unit" column of the Parameters
table on the "Unit" column of the Units table. 

```{r, eval=FALSE}
query <- dbSendQuery(db_par_unit, 
                     "SELECT Parameters.Parameter_Desc, 
                             Parameters.Standard_Unit,
                             Units.Unit_Desc
                      FROM Parameters
                      INNER JOIN Units
                      ON Parameters.Standard_Unit = Units.Unit")
param_units <- dbFetch(query)
head(param_units)
```
```{r}
param_units
```

Now we disconnect from the database and delete it.

```{r}
dbDisconnect(db_par_unit)
unlink("par_unit.sqlite")
```

[Back to exercises](#ex)