---
title: 'Data Manipulation: Part 2'
output: 
  html_document: 
    theme: readable
---

This tutorial will cover the following functions in the `dplyr` package:

- [mutate()](#mutate)

- [group_by() with filter()](#groupfilter)

- [group_by() with summarize()](#groupsum)

Advanced topic

- [Piping with %>%](#pipe)

## Make new columns with mutate() {#mutate}

In the first <a href="http://rpubs.com/NateByers/DataManip1" target="_blank">Data Manipulation</a> tutorial we learned how to select just the columns we need by using the `select()`
function from `dplyr`. But in some cases we may want to add a new column. For example, 
in the `chicago_air` data set from the `region5air` package, we may want a column 
that is a one day lag value of the ozone column. To add a new column we use the 
`mutate()` function from the `dplyr` package. Again, the first parameter is
the data frame. Then we use the `lag()` function to create a new column of lagged
ozone values.

```{r, warning=FALSE, message=FALSE}
library(region5air)
library(dplyr)
data(chicago_air)
head(chicago_air)
d <- mutate(chicago_air, ozone_lag1 = lag(ozone, n = 1))
head(d, 3)
```

We may want to also calculate the difference between the ozone value and a standard 
value.

```{r}
d <- mutate(d, ozone_lag1 = lag(ozone, n = 1), diff_standard = ozone - 0.075)
head(d)
```


## Filter by groups with group_by() and filter() {#groupfilter}

So far we have selected columns with `select()`, added new columns with 
`mutate()`, and filtered an entire data frame with `filter()`.

But sometimes when we filter we don't want to filter the entire data frame all at 
once. Suppose we are interested in the high values of ozone in the `chicago_air`
data frame. We can use the `max()` function to find the highest value.

```{r}
max(chicago_air$ozone, na.rm = TRUE)
```

But that's for the entire year. Suppose you wanted to know the maximum value for
each month. You could subset `chicago_air` 12 times and make a new data frame for
each month. Then find the maximum value for each data frame. But that would be 
very inefficient. What we really want to do is filter the data frame by first 
grouping all of the rows by month, then filtering on the highest ozone value in 
each month.

Below, a data frame named `ozone` is created for illustration. It contains ozone
values for one monitor, but for simplicity's sake it only has two days in June
and two days in July. 

```{r}
ozone <- read.table(header=T, text='
  day   month      value  unit                
    1    June      0.040   ppm
    2    June      0.050   ppm
    1    July      0.045   ppm
    2    July      0.052   ppm
                       ')
```

If we use the `filter()` function on `ozone` to get the maximum value, we'll get
one row.

```{r}
filter(ozone, value == max(value))
```

If we wanted to get the maximum value for each month, we would first use the `group_by()`
function.

```{r}
ozone <- group_by(ozone, month)
```

Now the `ozone` data frame is grouped by month. If we use the `filter()` function
now, we'll get two rows because there are two months.

```{r}
filter(ozone, value == max(value))
```

What if we had two columns we wanted to group by? Here is another simple data frame
to illustrate this situation.

```{r}
ozone_8h_1h <- read.table(header=T, text='
  day   month  duration    value  unit                
    1    June        1h    0.050   ppm
    1    June        8h    0.040   ppm
    2    June        1h    0.060   ppm
    2    June        8h    0.050   ppm
    1    July        1h    0.055   ppm
    1    July        8h    0.045   ppm
    2    July        1h    0.065   ppm
    2    July        8h    0.052   ppm
                       ')

```

Now we might want to find the maximum value by month _and_ by duration. We simply
add duration in the `group_by()` function.

```{r}
ozone_8h_1h <- group_by(ozone_8h_1h, month, duration)
```

Now the `filter()` function will return 4 rows, one for each month/duration pair.

```{r}
filter(ozone_8h_1h, value == max(value))
```

## Summarize by groups with group_by() and summarize() {#groupsum}

Filtering by grouped columns is appropriate for situations where we want to 
subset down to rows that meet a logical criteria. But sometimes we  want to apply
a summarizing function to grouped values. 

Let's say we want to find the mean for each month and duration in the 
`ozone_8h_1h` data frame. We've use the `group_by()` function already, 
`ozone_8h_1h <- group_by(ozone_8h_1h, month, value)`. We can't use `filter()` 
because there would be no value that satisfies the logical expression 
`value == mean(value)`. Instead, we use the `summarize()` function.

```{r}
summarize(ozone_8h_1h, mean = mean(value))
```

As you can see, the column "units" has been dropped. When you use `summarize()`
you will only keep the columns that have been grouped by and the column you create
with the summarizing function. All other columns are dropped.


### When should you use filter() and when should you use summarise()?

- __Use `filter()` with `group_by()`__ when you want to keep only the rows that 
meet a logical criteria among groups. You will be using `==` because it's a logical 
filter. The output will retain all of the columns.

- __Use `summarize()` with `group_by()`__ when you need a summary of one column
for each group. You will used `=` because you are creating a new column with a
summarizing function. The output will only retain the columns you grouped-by and
the one column you created with the summarizing function.

## Exercises

Go to the <a href="http://rpubs.com/NateByers/DataManip2Exercises" target="_blank">exercises page</a> for this tutorial and test your comprehension of this material.

-------------------------------------

## Advanced Topic

### Piping with %>% {#pipe}

`dplyr` takes advantage of the pipe operator, `%>%, from the `magrittr` package.

The `%>%` is a symbol you can use to chain `dplyr` functions together. This is 
convenient because it eliminates the necessity of creating new data frames, or 
re-assigning values to the same data frame.

Here's an example of a data manipulation process.

```{r}
data(airdata)

# filter down to ozone
ozone <- filter(airdata, parameter == "44201")

# select only certain columns
ozone <- select(ozone, site, datetime, duration, value, unit, poc)

# group by certain columns
ozone <- group_by(ozone, site, duration, poc)

# find mean value for each group
ozone <- summarize(ozone, mean = mean(value, na.rm = FALSE))

head(ozone)
```

With `%>%` you can start with the data frame and then write each `dplyr` function
without supplying the data frame.

```{r}
ozone <- airdata %>%
  filter(parameter == "44201") %>%
  select(site, datetime, duration, value, unit, poc) %>%
  group_by(site, duration, poc) %>%
  summarize(mean = mean(value, na.rm = FALSE))

head(ozone)
```