---
title: "Data Manipulation"
output: 
  html_document: 
    theme: readable
---

These exercises accompany the Data Manipulation Part 2 tutorial: 
<a href="http://rpubs.com/NateByers/DataManip2" target="_blank">
http://rpubs.com/NateByers/DataManip2</a>. These exercises use data frames
from the `region5air` library. Run the following code to clean out your global
environment and load the data you need:

```{r, message=FALSE}
rm(list = ls())
library(dplyr)
library(region5air)
data(airdata)
data(chicago_air)
```

## Exercises {#ex}

1. Use the `mutate()` function to add a column named "violation" to the 
`chicago_air` dataset that indicates when the ozone value was above 0.070 ppm. 
__Hint__: you will make the new column using a single equal sign, `violation =`.
The value you want is a logical comparison: `violation =` ozone values greater 
than 0.07.

[Solution 1](#ex1)

2. Use the `group_by()` function to group the `airdata` data frame by site, 
parameter, and duration. Then use `filter()` to make a data frame called `max_params`
with maximum values maximum values for each monitor/parameter/duration combination.
__Hint__: in `filter()` you will use `value ==`. You want to keep numbers in the
"value" column that equal the maximum for each group (`max()`, and don't forget
`na.rm = TRUE`).

[Solution 2](#ex2)

3. In exercise 2 we grouped the `airdata` data frame by site, parameter, and
duration. Use `summarize()` to create a data frame called `mean_params` with 
mean values for each monitor/parameter/duration combination. __Hint__: you will
use a single `=` symbol, `mean_value =`. You want to use the `mean()` function on 
numbers in the "value"" column for each group (don't forget `na.rm = TRUE`).

[Solution 3](#ex3)

----------------------------------

#### Advanced Exercise

4. Use the pipe, `%>%`, to string together this series of `dplyr` operations: 
subset `airdata` down to ozone (44201); group by site, datetime, and poc; use
`summarize` to find the mean value for each site/datetime combination; and filter
down to the top 4 values for each year. *Note*: this data set only
has 1 hour values. This exercise would make more sense with 8 hour data, for
comparison to the NAAQS.

[Solution 4](#ex4)




--------------------------------------

## Solutions

#### Solution 1{#ex1}

```{r}
chicago_air <- mutate(chicago_air, violation = ozone > 0.070)
head(chicago_air)
```

[Back to exercises](#ex)

#### Solution 2{#ex2}

```{r}
airdata <- group_by(airdata, site, parameter, duration)

max_params <- filter(airdata, value == max(value, na.rm = TRUE))
```


[Back to exercises](#ex)

#### Solution 3{#ex3}

```{r}
mean_params <- summarize(airdata, mean_value = mean(value, na.rm = TRUE))
```

[Back to exercises](#ex)

#### Solution 4{#ex4}

```{r}
ozone_4rth_high <- airdata %>%
  filter(parameter == "44201") %>%
  group_by(site, poc, datetime) %>%
  summarize(value = mean(value, na.rm = TRUE)) %>%
  mutate(year = substr(datetime, 1, 4)) %>%
  group_by(site, year) %>%
  filter(rank(value, ties.method = "max") == 4)
```


[Back to exercises](#ex)

