---
title: 'Data Manipulation Part 2: Exercises'
output: html_document
---

## Exercises {#ex}

These exercises accompany the [Data Manipulation 2](http://rpubs.com/NateByers/datamanip2) tutorial. 

1. Use the `mutate()` function to add two columns to the `chicago_air` dataset: a column named
"violation" indicating when the ozone value is above 0.075 ppm and a column named
"cumulative_total" that is the cumulative total of violations for the year. (*Hint: look at the `dplyr` vignette on window-functions.*) 

[Solution](#ex1)

2. Use the `gather()` function to reshape the `chicago_wind` data frame from a wide format
to a long format.

[Solution](#ex2)

3. Use the `spread()` function to turn the long data frame from exercise 2 back into
a wide data frame.

[Solution](#ex3)

4. Use the `summarize_each()` function to replace the `chicago_air` daily values with 
monthly maximum values.

[Solution](#ex4)

## Solutions

### Solution 1{#ex1}

Here's a good first stab at a solution.

```{r, message=FALSE, warning=FALSE}
library(region5air)
library(dplyr)

data(chicago_air)
ozone_violations <- mutate(chicago_air, violation = ozone > 0.075, 
                           cumulative_total = cumsum(violation))

```

But the `ozone_violations` data frame gives `NA` values in the "cumulative_total"
column at day 2013-07-01. This is because there are `NA` ozone values. We can solve
this by removing rows with ozone `NA`s first.

```{r}
ozone_violations <- filter(chicago_air, !is.na(ozone))
ozone_violations <- mutate(ozone_violations, 
                           violation = ozone > 0.075, 
                           cumulative_total = cumsum(violation))
head(ozone_violations, 3)
tail(ozone_violations, 3)
```

[Back to exercises](#ex)

### Solution 2 {#ex2}

```{r, message=FALSE, warning=FALSE}
library(tidyr)
data(chicago_wind)
head(chicago_wind, 3)
long_wind <- gather(chicago_wind, key = parameter, value = value,
                    wind_speed:ozone, na.rm = TRUE)
head(long_wind, 3)
```

[Back to exercises](#ex)

### Solution 3 {#ex3}

```{r}
wide_wind <- spread(long_wind, key = parameter, value = value)
head(wide_wind)
```

[Back to exercises](#ex)

### Solution 4 {#ex4}

```{r}
air_monthly <- group_by(chicago_air, month)
air_monthly <- summarize_each(air_monthly, funs(max(., na.rm = TRUE)),
                              -(weekday), -(date))
air_monthly
```

[Back to exercises](#ex)