---
title: "Plotting with `openair`"
output:
  html_document:
    theme: readable
  word_document: default
---

The `openair` package is specifically designed to plot air pollution data. This 
tutorial will give a brief introduction to many of the plotting functions
in `openair`. Visit the <a href="http://www.openair-project.org/" target="_blank">
project website</a> for more information, including
a comprehensive <a href="http://www.openair-project.org/Downloads/OpenAirManual.aspx" target="_blank">manual</a>.

This tutorial will cover the following `openair` functions.

- [`summaryPlot()`](#summary)

- [`windRose()`](#windrose)

- [`pollutionRose()`](#pollutionrose)

- [`timePlot()`](#timeplot)

- [`calendarPlot()`](#calendarplot)

## `summaryPlot()` {#summary}

The first plotting function we'll look at is `summaryPlot()`. Functions in the
`openair` package expect data frames that are in a certain format. The column with 
date and time information must be labeled "date" (lowercase) and be a `POSIXct` class.

The first data frame we'll use is the `chicago_air` data frame from the `region5air` 
package, which already has a column labeled "date".

```{r, message=FALSE, warning=FALSE}
library(region5air)
data(chicago_air)
head(chicago_air)
```

However, the class of that "date" column is `character`.

```{r}
class(chicago_air$date)
```

We need to change it using the `as.POSIXct()` function. This function will assume
that the format of the date is YYYY-MM-DD. So we just need to supply a correct time
zone.

```{r}
chicago_air$date <- as.POSIXct(chicago_air$date, tz = "America/Chicago")
```

If your date (and time) isn't in the default format, see the "Working with Dates
in R" section of <a href="http://rpubs.com/kfrost14/SubsetSort" target="_blank">this tutorial</a>. If you need to provide a different time zone, run the `OlsonNames()`
function for a list of options.

Now we just feed the first four columns of the data frame to the `summaryPlot()`
function. We use the `select()` function from `dplyr` to select our columns,
and we use the short hand `date:solar` to select the "date" column, the "solar"
column, and all the columns in between.

```{r, warning=FALSE, message=FALSE}
library(openair)
library(dplyr)

summaryPlot(select(chicago_air, date:solar))
```

The first column of the graph contains time series plots of all of the parameters
in the data frame. The red bars at the bottom of each panel show where there are
large segments of missing data. The panels on the right are the histograms of the
distributions for each parameter.

## `windRose()` {#windrose}

The `windRose()` function expects a data frame with columns for wind speed and wind 
direction labeled "ws" and "wd", respectively. Here we load the `chicago_wind` 
dataset from the `region5air` package and take a look at the columns.

```{r}
data(chicago_wind)
head(chicago_wind)
```

We need to create a "date" column with a `POSIXct` class, so we'll use the `as.POSIXct()`
function again. This time we will need to provide information in the `format`
parameter.

```{r}
chicago_wind$datetime <- as.POSIXct(chicago_wind$datetime, format ="%Y%m%dT%H%M",
                                    tz = "America/Chicago")
```

Now we'll rename the columns using the `rename()` function from `dplyr`.

```{r}
chicago_wind <- rename(chicago_wind, date = datetime, ws = wind_speed, wd = wind_dir)
```

Then we can feed the data frame to the `windRose()` function.

```{r}
windRose(chicago_wind, key.footer = "knots") # default is m/s
```

We can split the data frame by time periods by using the `type` argument.

```{r, warning=FALSE}
windRose(chicago_wind, type = "weekday", key.footer = "knots")
```

## `pollutionRose()` {#pollutionrose}

We can make a similar plot that will display pollutant concentrations in relation 
to wind direction.

```{r}
pollutionRose(chicago_wind, pollutant = "ozone")
```

We can also look at the values by time periods.

```{r, warning=FALSE}
pollutionRose(chicago_wind, pollutant = "ozone", type = "month")
```

## `timePlot()` {#timeplot}

Time series plots can be easily produced using `timePlot()`

```{r, warning=FALSE}
timePlot(chicago_air, pollutant = c("ozone", "temp", "solar"))
```


## `calendarPlot()` {#calendarplot}

`calendarPlot()` displays daily values in a calendar format.

```{r, warning=FALSE}
calendarPlot(chicago_air, pollutant = "ozone")

calendarPlot(chicago_wind, pollutant = "ozone", annotate = "ws")
```

## Exercises

Exercises for this tutorial can be found here: 
<a href="http://rpubs.com/NateByers/OpenairExercises" target="_blank">
http://rpubs.com/NateByers/OpenairExercises</a>.