---
title: "Plotting with `openair`"
output: html_document
---

The `openair` package is specifically designed to plot air pollution monitoring
data. This tutorial will give a bried introduction to many of the plotting functions
in `openair`.

- [summaryPlot](#summary)
- [windRose](#windrose)
- [pollutionRose](#pollutionrose)
- [timePlot](#timeplot)
- [calendarPlot](#calendarplot)

# `summaryPlot` {#summary}

The first plotting function we'll look at is `summaryPlot()`. Functions in the
`openair` package expect data frames that are in a certain format. The column with 
time information must be labeled `date` (lowercase) and be a `POSIX` class.

```{r, message=FALSE, warning=FALSE}
library(region5air)
data(chicago_air)
head(chicago_air, 3)
chicago_air$date <- as.POSIXct(chicago_air$date, format = "%Y-%m-%d")

library(openair)
summaryPlot(chicago_air[, 1:4])
```

The first column of the graph contains time series plots of all of the columns in 
the data frame (besides `date`). The red bars at the bottom of each panel shows 
where there are large segments of missing data. The panels on the right are the histograms
of the distributions for each parameter.

# `windRose`{#windrose}

The `windRose()` function expects a data frame with columns for wind speed and wind direction
labeled `ws` and `wd`, respectively. Here we load the `chicago_wind` dataset and take a look 
at the columns.

```{r}
data(chicago_wind)
head(chicago_wind, 3)
```

We need to create a `date` column with a `POSIX` class and rename the wind columns before
feeding the data frame to `windRose()`.

```{r}
date <- substr(chicago_wind$datetime, 1, 13)
date <- as.POSIXct(date, format = "%Y%m%dT%H%M")
chicago_wind$datetime <- date
names(chicago_wind)[1:3] <- c("date", "ws", "wd")

windRose(chicago_wind, key.footer = "knots") # default is m/s
```

You can split the data frame by time periods by using the `type` argument.

```{r, warning=FALSE}
windRose(chicago_wind, type = "weekday", key.footer = "knots")
```

# `pollutionRose`{#pollutionrose}

You can make a similar plot that will display pollutant concentrations in relation to wind
direction.

```{r}
pollutionRose(chicago_wind, pollutant = "ozone")
```

You can also look at the values by time periods.

```{r, warning=FALSE}
pollutionRose(chicago_wind, pollutant = "ozone", type = "month")
```

# `timePlot`{#timeplot}

Time series plots can be easily produced using `timePlot()`

```{r, warning=FALSE}
timePlot(chicago_air, pollutant = c("ozone", "temp", "solar"))
```

There is an option to normalize all of the parameters and put them all in one plot.

```{r, warning=FALSE}
timePlot(chicago_air, pollutant = c("ozone", "temp", "solar"),
         avg.time = "month", normalise = "1/1/2013", lwd = 4, lty = 1,
         group = TRUE)
```

# `calendarPlot`{#calendarplot}

`calendarPlot()` displays daily values in a calendar format.

```{r, warning=FALSE}
calendarPlot(chicago_air, pollutant = "ozone")

calendarPlot(chicago_wind, pollutant = "ozone", annotate = "ws")
```