---
title: "Functions and the apply() Family"
output: html_document
---

- [Writing functions](#functions)
- [For loop](#loops)
- [apply() and related functions](#apply)

If we can do _anything_ in R, then we already know how to use functions. This tutorial
will cover how to create our own functions. We'll also review some looping functions,
including the infamous `apply()` family of functions.


# Writing functions {#functions}

We already know that when we use a function, we need to know what the arguments are.
For example, if we want to use the `mean()` function, we look at the documentation
by typing `?mean()`.

The first function we write will be our own version of a funciton that calculates
the mean of a numeric vector. Since we can't just call it `mean` (or, if we do, we will
replace that base function in our environment), we will call our function `myMean`.

Here is the basic structure of creating a function.

```{r, eval=FALSE}
myMean <- function( ){
  
}
```

A function is created by using another function called... `function()`. The parentheses will
contain the parameters we want our function to have, and the curly braces will contain the 
operation that will be performed on the parameters. First, we'll need a vector of numbers,
so we'll call it the `x` parameter.

```{r, eval=FALSE}
myMean <- function(x){
  #do something to x
}
```

Now we'll calculate the average of the numbers in `x`.

```{r, eval=FALSE}
myMean <- function(x){
total_count_of_values <- length(x)
total_sum_of_values <- sum(x)
average_of_values <- total_sum_of_values/total_count_of_values
average_of_values
}
```

We've found how many numbers are in the `x` vector, we've added up all the values
in that vector, and we've found the average. In a function, the last line will always 
be returned. You can also use the `return()` function, but it's not necessary.

```{r}
myMean <- function(x){
total_count_of_values <- length(x)
total_sum_of_values <- sum(x)
average_of_values <- total_sum_of_values/total_count_of_values
return(average_of_values)
}
```

Let's try our function.

```{r}
my_vector <- c(1, 3, 5, 2, 6, 9, 0)
vector_mean <- myMean(x = my_vector)
vector_mean
```

# For loop {#loops}

Like most programming languges, R has for and while loops. We'll just review for loops
and move on to `apply()` functions, which are more commonly used in R.

For loops are used to repeat an operation a set number of times. The basic outline is

```{r, eval=FALSE}
for(i in sequence){
  
}
```

The `sequence` parameter is typically a vector. The `i` parameter is a variable that 
will take on the values in the `sequence` vector. For instance, if `sequence` was the vector `c(1, 2, 3)` then the `i` will take on each of those values in turn.

Here we use our `myMean()` function to find the average of three vectors.

```{r}
my_list <- list(c(1, 5, 9, 3), 1:10, c(23, 42))
my_averages <- c()
for(i in c(1, 2, 3)){
  my_averages[i] <- myMean(my_list[[i]])
}
my_averages
```

We use the `i` parameter as a variable to extract different members of `my_list` and
add the average to different positions in the `my_averages` vector.

We could use the `i` parameter to specify a column in a data frame. Let's use a for
loop to calculate the means of a few columns in the `chicago_air` dataset from the 
`region5air` package.

```{r}
library(region5air)
data(chicago_air)
head(chicago_air)

chicago_avgs <- c()
for(i in c("ozone", "temp", "solar")){
  chicago_avgs[i] <- myMean(chicago_air[, i])
}
chicago_avgs
```

It looks like there is a problem with `myMean()`. We didn't account for `NA`s. A
properly written function would need to take into account that sort of thing, but
we can also deal with `NA`s in the for loop.

```{r}
chicago_avgs <- c()
for(i in c("ozone", "temp", "solar")){
  numeric_series <- chicago_air[, i]
  numeric_series <- numeric_series[!is.na(numeric_series)]
  chicago_avgs[i] <- myMean(numeric_series)
}
chicago_avgs
```

# The apply() family {#apply}

In R, the most efficient way to do loops is to use the `apply()` functions. These are
functions that have `apply()` at the end of their name (such as `lapply()`, `tapply()`, 
and `mapply()`) and apply functions to each member of a vector, list, or column in a 
data frame. 

- `apply()` takes a data frame (or matrix) as the first argument. The second argument 
specifies if you want to apply a function to the rows (`1`) or columns (`2`), and the 
third argument is the function you want to apply to each row or column. Additional 
arguments can be used to pass on to the function being applied to each row or column.

```{r}
air <- chicago_air[, c("ozone", "temp", "solar")]


air_max <- apply(air, 
                 MARGIN = 2, # we are applying the max() function to each column
                 FUN = max, 
                 na.rm = TRUE# na.rm is being passed to the max() function
                 )
air_max
```

- `lapply()` applies a function to each member of a list. Here we find the length
of each vector in `my_list`.

```{r}
lapply(my_list, length)
```

- `sapply()` does the same thing as `lapply()` but the output is simplified as much
as possible. Also, for any of the `apply()` functions, the `FUN` argument can be
specified inside the function itself. Here we find how many `NA`s are in each column
of the `air` data frame. (Since a data frame is really a list of columns, `air` can
be used as the list argument.)

```{r}
sapply(air, function(column){
  the_NAs <- is.na(column) # returns a logical vector, TRUE if NA
  sum(the_NAs) # since TRUE is equivalent, this gives a count of NAs
  })
```

- `tapply()` takes (typically) a vector of values that a function will be applied to.
The second argument is a list of one or more factors that will split the first vector
into sections that the function will then be applied to. The third argument is the
function, and additional arguments can be passed on to the function. Here we find the max
ozone values in the `chicago_air` data frame by month.

```{r}
tapply(chicago_air$ozone, list(chicago_air$month), max, na.rm = TRUE)
```
